---
title: "Analysis of impact of weather events in population health and local economy"
author: "Carlos Figueira"
date: "Wednesday, March 18, 2015"
output: 
  html_document:
    keep_md: true
---

The National Weather Service has tracked major storms and weather events over more than 50 years in the United States, creating a catalog of all such events and their consequences to the population around where they occurred. In this project we'll analyze the National Oceanic and Atmospheric Administration's (NOAA) [Storm Events Database](http://www.ncdc.noaa.gov/stormevents/details.jsp), and their impact in the health of the population and economy of affected areas. The database for this analysis was obtained from [the address in the Coursera Reproducible Research class page](http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) instead of the NOAA directly.

## Data Processing

Before we begin processing data, let's load the dataset, downloading it from the source if it hasn't already been downloaded. At this point we'll also load the libraries that we'll use throughout this document.

```{r echo=FALSE}
# Load required libraries
library(dplyr)
library(ggplot2)
```

```{r echo=TRUE, cache=TRUE}
# Download dataset if it doesn't exist
remoteCompressed <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
localCompressed <- "StormData.csv.bz2"
datasetFolder <- "dataset"

compressedDatasetFile <- paste(datasetFolder, localCompressed, sep = "/")

if (!file.exists(datasetFolder)) {
  dir.create(datasetFolder)
}

if (!file.exists(compressedDatasetFile)) {
    download.file(remoteCompressed, destfile = compressedDatasetFile, mode = "wb", method = "auto")
}

# Load the dataset to memory
dataset <- read.csv(bzfile(compressedDatasetFile))
```

Having the data loaded, we'll need to clean it up a little, by removing the variables that we don't need for this analysis. We'll also to tidy up the data, merging the crop/property damage value and exponent into a single column.

```{r echo=TRUE, cache=TRUE}
stormData <- select(dataset, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, BGN_DATE)
rm(dataset) # Remove dataset from memory, not needed

# Extract year from the date
eventYears <- sapply(
  strsplit(
    as.character(stormData$BGN_DATE),
    "[\\/ ]"),
  function(x) { as.integer(x[3]) })
stormData$year = eventYears
stormData$BGN_DATE <- NULL

# Normalize property / crop damages to millions of dollars
normalizeValue <- function(base, exp) {
  exp <- tolower(as.character(exp))
  exp <- sapply(tolower(as.character(exp)), function(exp) {
    if (length(grep("\\d", exp)) > 0) {
        exponent <- as.numeric(exp)
    } else if (exp == "h") { # hundreds
        exponent <- 2
    } else if (exp == "k") { # thousands
        exponent <- 3
    } else if (exp == "m") { # millions
        exponent <- 6
    } else if (exp == "b") { # billions
        exponent <- 9
    } else {
        exponent <- 0
    }
    exponent
  })
  base * 10 ^ (exp - 6)
}

# Clean up the data, and filter only those that have an impact in the variables
# that we want to analyze (i.e., remove those events that did not cause any
# fatalities, injuries or economic damages)
tidyStormData <-
  stormData %>%
  filter(FATALITIES > 0 | INJURIES > 0 | CROPDMG > 0 | PROPDMG > 0) %>%
  mutate(evtType = tolower(as.character(EVTYPE)), propertyDamage = normalizeValue(PROPDMG, PROPDMGEXP), cropDamage = normalizeValue(CROPDMG, CROPDMGEXP)) %>%
  select(year, evtType, fatalities = FATALITIES, injuries = INJURIES, propertyDamage, cropDamage)
```

To continue cleaning up the data, we'll look at the event type. As we can see below, there are too many event types, and many of them represent the same type (such as "winter weather mix" and "winter weather/mix").

```{r echo=TRUE}
length(table(tidyStormData$evtType))
```

Let's normalize them into some groups to better capture their effect that we want to analyze.

```{r echo=TRUE}
assignGroup <- function(groupName, options) {
    regex <- paste(options, collapse = ")|(")
    regex <- paste0("(", regex, ")")
    tidyStormData[grep(regex, tidyStormData$evtType), "evtGroup"] <<- groupName
}

# Create groups of events based on their types:
tidyStormData$evtGroup <- "others"
assignGroup("rain / storms",
            c("thunderstorm", "tstm wind", "lightning", "(strong)|(high) wind",
              "tropical storm", "heavy rain", "hurricane", "waterspout", "storm surge",
              "landslide", "high surf", "heavy surf", "tsunami", "wind", "typhoon",
              "excessive rainfall", "excessive wetness", "mixed precip"))
assignGroup("tornado / hail",
            c("tornado", "hail", "funnel", "sleet"))
assignGroup("flood",
            c("flood", "stream fld", "rip current"))
assignGroup("winter",
            c("snow", "blizzard", "avalanche", "winter", "ice", "freez", "cold",
              "ice", "frost", "icy", "windchill", "hypothermia", "wintry", "glaze",
              "unusual cool"))
assignGroup("summer / heat",
            c("fire", "heat", "drought", "dust storm", "dust devil",
              "dry microburst", "hyperhermia", "unseasonably warm",
              "record warmth", "unseasonably ((dry)|(hot))", "unusual warm"))
assignGroup("fog", c("fog", "smoke"))
```

We're still left with some events without a categorization, but their number and the impact is minimal for this analysis.

```{r echo=TRUE}
table(tidyStormData$evtGroup)
```

Now we can try to understand the impact of those event types in the local economy / population health. Let's start with harm to the population health. We'll include both injuries and fatalities related to the events, summarizing the data set to aggregate the total number of injuries / fatalities caused by them. Also, since some groups have a lot more events than others (there are a lot more storms), we'll also look at the impact caused by an average event.

```{r echo=TRUE, fig.width=9}
healthImpact <- tidyStormData %>%
    group_by(evtGroup) %>%
    summarise(count = n(),
              avgInjuries = mean(injuries),
              avgFatalities = mean(fatalities),
              injuries = sum(injuries),
              fatalities = sum(fatalities))
print(healthImpact)

# Mutate the data to make it more amenable to ggplot
healthImpactForPlot <- rbind(
    tidyStormData %>%
        group_by(evtGroup) %>%
        summarise(eventsPerGroup = n(),
                  impactType = "injuries",
                  aggregation = "total",
                  value = sum(injuries)),
    tidyStormData %>%
        group_by(evtGroup) %>%
        summarise(eventsPerGroup = n(),
                  impactType = "injuries",
                  aggregation = "average",
                  value = mean(injuries)),
    tidyStormData %>%
        group_by(evtGroup) %>%
        summarise(eventsPerGroup = n(),
                  impactType = "fatalities",
                  aggregation = "total",
                  value = sum(fatalities)),
    tidyStormData %>%
        group_by(evtGroup) %>%
        summarise(eventsPerGroup = n(),
                  impactType = "fatalities",
                  aggregation = "average",
                  value = mean(fatalities)))

ggplot(healthImpactForPlot, aes(x = evtGroup, y = value, fill = impactType)) +
    geom_bar(stat = "identity") +
    facet_grid(aggregation ~ ., scales = "free") +
    xlab("Type of event") +
    ylab("Population health impact (number of occurrences)")
```

Now for the impact in the local economy. The analysis is similar to the one above, with respect to property and crop damages:

```{r echo=TRUE, fig.width=9}
economicImpact <- tidyStormData %>%
    group_by(evtGroup) %>%
    summarise(count = n(),
              property = sum(propertyDamage),
              avgProperty = mean(propertyDamage),
              crop = sum(cropDamage),
              avgCrop = mean(cropDamage))
print(economicImpact)

# Mutate the data to make it more amenable to ggplot
economicImpactForPlot <- rbind(
    tidyStormData %>%
        group_by(evtGroup) %>%
        summarise(eventsPerGroup = n(),
                  impactType = "property",
                  aggregation = "total",
                  value = sum(propertyDamage)),
    tidyStormData %>%
        group_by(evtGroup) %>%
        summarise(eventsPerGroup = n(),
                  impactType = "property",
                  aggregation = "average",
                  value = mean(propertyDamage)),
    tidyStormData %>%
        group_by(evtGroup) %>%
        summarise(eventsPerGroup = n(),
                  impactType = "crop",
                  aggregation = "total",
                  value = sum(cropDamage)),
    tidyStormData %>%
        group_by(evtGroup) %>%
        summarise(eventsPerGroup = n(),
                  impactType = "crop",
                  aggregation = "average",
                  value = mean(cropDamage)))

ggplot(economicImpactForPlot, aes(x = evtGroup, y = value, fill = impactType)) +
    geom_bar(stat = "identity") +
    facet_grid(aggregation ~ ., scales = "free") +
    xlab("Type of event") +
    ylab("Economic impact (millions of dollars)")
```

## Results

From the numbers and plots above, we can see that the tornado / hail events are the ones that have caused the larges number of injuries / fatalities. However, on a per-case basis, dense fog and extreme heat events have a larger impact on the population health, with the larger number of fatalities happening on extreme heat events.

The economic impact of the various event types is different. Looking at this side, floods and storms are the types of events that have caused the largest economic impact, with over 330 billions USD worth of property damages and over 20 billion USD of crop damages.over the years. On a per-case basis, the worst type of recorded events are summer / heat (such as droughts) that cause in average close to 9 million USD worth of damages, most of it in crop damages.
